---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# txtools

<!-- badges: start -->
`r badger::badge_devel("AngelCampos/txtools", "blue")`
<!-- badges: end -->

## Description 

**txtools** is a package that processes genome aligned RNA-seq data into 
transcriptomic-oriented data.tables (DT), enabling a quick and simplified 
analysis, as well as close inspection of summary metrics per transcript, at 
nucleotide resolution. A quick example of its capabilities is shown bellow.

## Quick example

As an example we will use the 'Pasilla' experiment data (from its own package)
which contains a BAM file for the paired-end alignments of a *D. melanogaster* 
RNA-seq experiment on chromosome 4, along with a FASTA file comprising the 
genome sequence for the same chromosome.

Using txtools we can load the **genome** (FASTA), the **gene annotation** 
(BED-12), and the **RNA-seq reads alignment** (BAM) files into R with ease.

```{r, results = "hide"}
# Load packages
library(txtools)
library(pasillaBamSubset)

# Getting paths to files
BED_file <- tx_dm3_geneAnnot()
FASTA_file <- dm3_chr4()
PE_BAM_file <- untreated3_chr4()

# Loading gene annotation, genome, and alignments into R.
dm3_geneAnnot <- tx_load_bed(BED_file)
dm3_genome <- tx_load_genome(FASTA_file)
dm3_PEreads <- tx_load_bam(file = PE_BAM_file, pairedEnd = T, loadSeq = T)
```

Then we process the alignments to their transcriptomic versions using the
`tx_reads()` function, and summarize the alignments into a DT that contains 
coverage, and nucleotide frequency count data using the 
`tx_makeDT_covNucFreq()`function.

```{r}
reads_SE <- tx_reads(reads = dm3_PEreads, 
                     geneAnnot = dm3_geneAnnot, 
                     withSeq = T, 
                     nCores = 10, 
                     minReads = 1)

DT <- tx_makeDT_covNucFreq(reads_SE, geneAnnot = dm3_geneAnnot, genome = dm3_genome)
```

The resulting DT comprise all summarized information from the RNA-seq reads 
aligned to the genome and contained within the genes in the gene annotation 
(this example consists of only the top 10 expressed genes). For more information on the
columns of DT consult the `tx_makeDT_covNucFreq()` documentation.

To extend the base metrics that the `tx_makeDT_*()`
functions provide txtools provides the `tx_add_*()` functions family. One
example of such functions is `tx_add_diffNucToRefRatio()` which calculates the 
ratio of nucleotide counts different to the reference sequence. Using
these metric we can easily spot locations in which RNA transcripts sequence is
different from that of the reference sequence. 

```{r}
DT <- tx_add_diffNucToRefRatio(DT, addDiffandTotalCols = TRUE)
DT[which(diffToRefRatio > 0.5 & nucTotal > 40),]
```

Finally, using the `tx_plot_nucFreq()` 
function we can visualize that data in the DT at an specific location.

```{r, fig.width = 6, fig.height = 4}
tx_plot_nucFreq(DT, gene = "NM_079901", txRange = window_around(3803, 15))
```

## Installation

You can install the development version from
[GitHub](https://github.com/AngelCampos/txtools) typing in the following 
commands in the R console:

```{r, installation, eval = F}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("AngelCampos/txtools")
```

## Further documentation
    
## Current limitations:

* Insertions: txtools is not able to deal with insertions. This is mainly
because insertions are not part of the original trasncriptomic nor genomic 
reference space as they would alter the length of the gene model. 
This could be an added feature in future versions but is not a priority. 

* Potentially long processing times: Loading big BAM files into R commonly 
requires a lot of time, having this in mind txtools provides a progress bar to 
keep users informed about the loading status. Most importantly, depending on the
ammount of both loaded reads and the size of the *Gene Annotation* tx_reads() 
processing time can take several minutes. A solution to this issue is the use of
multi-threadding which has been incorporated into tx_reads() and other 
functions, but such functionality is only available for UNIX and MAC OS.

## Additional notes:

* As many R packages meant for high-throughput data analysis and manipulation, 
using ***txtools*** may require high ammounts of RAM memory, depending mainly on
the size of BAM files being processed.

## Session Info
```{r}
utils::sessionInfo()
```
